---
title: "Incremental refresh of star databases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Incremental refresh of star databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message=FALSE}
library(rolap)
```

# Introduction

Once we have star databases built with the data available at the moment, periodically we may obtain additional data, with the same structure as the initial data but from a later time or from another place. Sometimes the new data also contains instances which are already included in the star database to operate on them.

Suppose we need to jointly analyse all the data available at the moment: We must include them in the star database. One possibility is to integrate all available data into a flat table and build the star database again. Another possibility is to use the *incremental refresh* mechanism described in this document.

This document shows by means of an example the possibilities offered by the package in this context. First,  , then  . Finally, the conclusions are presented.

# Starting data set

The starting data set is the content in the variable `mrs_db`, obtained in the vignette titled *Obtaining and transforming flat tables*, `vignette("v05-flat-table-op")`.


# Conclusions

The `rolap` package supports the definition
